Linear response theory for a system under non-equilibrium conditions

### 2013-06-28 writedown/linear.response/paper.tex
update changes

\usepackage[fleqn]{amsmath}
\usepackage{amssymb}
\usepackage[dvips]{graphicx}
\usepackage{color}
\usepackage{tabularx}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

@@

\newcommand{\mh}{\mathcal H}
\newcommand{\eps}{\varepsilon}
\newcommand{\ml}{\mathcal L}
\newcommand{\mt}{\mathcal T}
\newcommand{\mo}{\mathcal O}
\newcommand{\mi}{\mathcal I}
\newcommand{\mc}{\mathcal C}
\newcommand{\proj}{\mathit\Pi}
\newcommand{\fwg}{{\mathcal A}}
\newcommand{\bwg}{{\mathcal B}}
\newcommand{\bsigma}{\boldsymbol\sigma}

@@

% \title{Linear response theory for core set identification}
\title{The response theory for the system under a non-equilibrium driving}
\author{Han Wang}
\author{Carsten Hartmann}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\affiliation{Zuse Institute Berlin (ZIB), Germany}
% \affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}

@@

\begin{abstract}
  In this talk, we firstly review the traditional linear response
  theory, and then generalize it to the system under a non-equilibrium
  driving. We calculate not only the linear response term, but also
  the higher order terms with respect to the perturbation to the
  reference dynamics. The response theory is applied to investigate
  the dynamics of core set that plays a central role in the Markov
  State Model (MSM). The advantages and limitations are discussed from
  the numerical tests.
  % In this draft, we are trying to study the core set based Markov
  % State Model (MSM) under non-equilirbium conditions.
\end{abstract}

@@

\section{Theoretical background}
\subsection{Classical equilibrium linear response theory}

In this section, we want to firstly give a brief description of the classical linear response
theory, for details see Ref.~\cite{tuckeman2010statistical} for
example.
We study the non-equilibrium system driven by some external force.
More strictly, we investigate the system that is governed by the
following Langevin equation:
\begin{align}\label{eqn:langevin-1}
  \dot{\vect q} & = \nabla_{\vect p}\mh(\vect q,\vect p), \\\label{eqn:langevin-2}
  \dot{\vect p} & =- \nabla_{\vect q}\mh(\vect q,\vect p)
  + \eps F_e(t) \vect D(\vect q) 
  - \gamma\vect p
  + \sigma\dot{\vect W}.
\end{align}
Here $\vect D$ is the external driving force applied to the system,
the magnitude of which is always assumed to be controllable.
$\eps F_e(t)$ is the strength of the driving as a function of time.
The prefactor $\eps$ indicates that we regard the driving force to be considerably smaller than the internal forces. We will comment on what is meant by "considerably smaller" further below.

% , and
% the magnitude of $F_e(t)$ is small and of order $\mo (\eps)$, $\eps \ll 1$.

Then the phase space distribution $f(\vect x, t)$ ($\vect x = \{\vect
q, \vect p\}$) is subject to the Kolmogorov forward equation:
\begin{align}\label{eqn:orig-forward}
  \frac{\partial}{\partial t} f(\vect x, t) - \fwg^\beta(t) f(\vect x, t) = 0
\end{align}
where $\fwg^\beta(t)$ is the forward infiniesimal generator given by
\begin{align}
  \fwg^\beta(t) =
  \frac{\sigma^2}2\Delta_{\vect p}
  - \nabla_{\vect p}\mh 
  \cdot\nabla_{\vect q}
  - \Big(
  -\nabla_{\vect q}\mh +
  \eps F_e(t)  \vect D  - \gamma\vect p
  \Big)\cdot\nabla_{\vect p}
  + 3N\gamma,
\end{align}
and $\beta2=\gamma / \sigma^2$ is the inverse temperature.
The equilibrium system
is governed by
the standard Langevin equation, i.e. with
vanishing driving force $\eps = 0$ in
Eqn.~\eqref{eqn:langevin-1} and \eqref{eqn:langevin-2}.

Under our smallness assumption on the driving force the full phase space distribution has the following asymptotic expansion in $\eps$:
\begin{align}\label{eqn:f-expan}
  f(\vect x, t) &= f_0(\vect x, t) + \eps f_1(\vect x, t)
  +\eps^2 f_2(\vect x, t) + \mo (\eps^3),
\end{align}
where $f_0=f_0(x,t)$ denotes the evolution of the phase space distribution for the undriven system (with $\eps=0$).
In addition we have:
\begin{align}\label{eqn:A-expan}
  \fwg^\beta(t) = \fwg^\beta_0 + \eps\fwg_1(t),
\end{align}
where
\begin{align}
  \fwg_0^\beta =&
  \frac{\sigma^2}2\Delta_{\vect p}
  -
  \nabla_{\vect p}\mh\cdot\nabla_{\vect q}
  - \big(
  -\nabla_{\vect q}\mh - \gamma\vect p
  \big)\cdot\nabla_{\vect p}
  + 3N\gamma,\\
  \fwg_1(t) =&
  - F_e(t)  \vect D \cdot\nabla_{\vect p}.
\end{align}

The invariant measure of $\fwg^\beta_0$ satisfies $\fwg^\beta_0f^\beta_{\inv}(\vect x) = 0$, and is given by the Boltzmann distribution
\begin{align}
  f^\beta_{\inv}(\vect x) \propto e^{-\beta\mh(\vect x)}
\end{align}
Inserting \eqref{eqn:f-expan} and \eqref{eqn:A-expan} into
\eqref{eqn:orig-forward}, 
the Kolmogorov forward equation becomes:
\begin{align}
  \frac{\partial}{\partial t}
  [f_0(\vect x, t) + \eps  f_1(\vect x, t) + \eps^2f_2(\vect x, t)]
  -
  [\,\fwg^\beta_0 + \eps\fwg_1(t)\,]
  [f_0(\vect x, t) + \eps  f_1(\vect x, t) + \eps^2f_2(\vect x, t)]
  = 0
\end{align}
Matching this equation at different orders of $\eps$, the zeroth
order gives:
\begin{align}\label{eqn:e-o0}
  \bigg[
  \frac{\partial}{\partial t}
  - \fwg_0^\beta
  \bigg]
  f_0(\vect x, t)
  = 0
\end{align}
which is the Kolmogorov forward equation of the undriven system. Matching the first and the second order gives:
\begin{align}\label{eqn:e-o1}
  \bigg[
  \frac{\partial}{\partial t}
  - \fwg_0^\beta
  \bigg]
  f_1(\vect x, t)
  =&
  \fwg_1(t) f_0(\vect x, t)\\\label{eqn:e-o2}
  \bigg[
  \frac{\partial}{\partial t}
  - \fwg_0^\beta
  \bigg]
  f_2(\vect x, t)
  =&
  \fwg_1(t) f_1(\vect x, t)
\end{align}
Formally solving \eqref{eqn:e-o0} -- \eqref{eqn:e-o2} with
initial conditions $f_1(\vect x, 0) = 0$ and $f_2(\vect x, 0) = 0$
gives
\begin{align}\label{eqn:f0-0}
  f_0(\vect x, t)
  =&\,
  e^{t\fwg^\beta_0}f_0(\vect x, 0) \\\label{eqn:f1-0}
  f_1(\vect x, t)
  =&\,
  \int_0^t\dd s\,
  e^{(t-s)\fwg^\beta_0}\circ
  \fwg_1(s) f_0(\vect x,s)
  =
  \int_0^t\dd s\,
  e^{(t-s)\fwg^\beta_0}\circ
  \fwg_1(s)\circ
  e^{s\fwg^\beta_0}f_0(\vect x, 0) \\\nonumber
  f_2(\vect x, t)
  =&\,
  \int_0^t\dd s\,
  e^{(t-s)\fwg^\beta_0}\circ
  \fwg_1(s) f_1(\vect x,s) \\\label{eqn:f2-0}
  &=
  \int_0^t\dd s
  \int_0^s\dd u\,\,
  e^{(t-s)\fwg^\beta_0}\circ
  \fwg_1(s)\circ
  e^{(s-u)\fwg^\beta_0}\circ
  \fwg_1(u)\circ
  e^{u\fwg^\beta_0}
  f_0(\vect x, 0) 
\end{align}
If the initial distribution is the invariant distribution, i.e.
$f_0(\vect x, 0) = f_{\inv}^\beta(\vect x)$,
\eqref{eqn:f0-0} becomes:
\begin{align}
  f_0(\vect x, t) = f_{\inv}^\beta(\vect x)
\end{align}
It can be shown that
\begin{align}
  \fwg_1(s) f_{\inv}^\beta(\vect x)
  =
  -\beta {F_e(s)}
  j(\vect x)
  f_{\inv}^\beta(\vect x),
\end{align}
where the \emph{dissipative flux} $j(\vect x)$ is defined by,
\begin{align}
  j(\vect x) =
  -\vect D\cdot\nabla_{\vect p}\mh
\end{align}
Then $f_1(\vect x, t)$ becomes:
\begin{align}
  f_1(\vect x, t)
  =
  -\beta
  \int_0^t\dd s\,
  F_e(s)\,
  e^{(t-s)\fwg^\beta_0}
  \big[
  j(\vect x)
  f_{\inv}^\beta(\vect x)
  \big]
\end{align}
Therefore, the linear order approximation to
the solution of~\eqref{eqn:orig-forward} is 
\begin{align}\label{eqn:solv-orig-linear}
  f(\vect x, t) =
  f_{\inv}^\beta(\vect x)
  - \eps\beta
  \int_0^t\dd s\,
  F_e(s)\,
  e^{(t-s)\fwg^\beta_0}
  \big[
  j(\vect x)
  f_{\inv}^\beta(\vect x)
  \big]
  + \mo(\eps^2)
\end{align}
Now let $\mathcal{O}$ be an observable of the system. Its expectation value satisfies the linear response formula:
\begin{align}\nonumber
  \mathcal O(t)
  &=
  \int\dd \vect x \:O(\vect x)f(\vect x, t)  \\\nonumber
  &=
  \int\dd \vect x\, O(\vect x)\,
  [\,f_{\inv}^\beta(\vect x) + \eps f_1(\vect x, t)] +\mathcal{O}(\eps^2)\\\nonumber
   \label{eqn:eqi-pert-1}
  &=
  \langle O\rangle
  -
 \eps \beta
  \int_0^t\dd s\,
  F_e(t-s)
  \int \dd \vect x\,
  f^\beta_{\inv}(\vect x) j(\vect x)\,
  e^{s\bwg_0^\beta}
  O(\vect x)+\mathcal{O}(\eps^2)
\end{align}
where $ \langle O\rangle=\int \dd\vect x\,O(\vect x) f_{\inv}^\beta(\vect x)$ is the expectation value of the undriven equilibrium system, and $\bwg^\beta_0$ is the infiniesimal generator of the undriven 
Kolmogorov backward equation, which is the adjoint operator of $\fwg^\beta_0$. That is,
we have
\begin{align}
  e^{s\bwg^\beta_0} O(\vect x) = \mathbb E_{\vect x} [O(\vect X^\beta_s)]
\end{align}
where $\vect X^\beta_t$ is the trajectory of the undriven Langevin equation (with $\eps=0$)
starting at $\vect x$.
Therefore, we have up to order 2 in $\eps$:
% Notice that $e^{i\ml_0(t-s)}O(\vect x) = O(\phi_{t-s}(\vect x))$,
% where $\phi_t(\vect x)$ is the flow mapping of the \emph{unperturbed}
% Hamiltonian system. We have
\begin{align}\nonumber
  \mathcal O(t)
 & =
  \langle O\rangle
  -
  \eps\beta
  \int_0^t\dd s\,
  F_e(t-s)
  \int \dd \vect x\,
  f^\beta_{\inv}(\vect x)
  j(\vect x)\,
  \mathbb E_{\vect x} [O(\vect X^\beta_s)]\\
\label{eqn:eq-lr}  
 & =
  \langle O\rangle
  -
  \eps \beta
  \int_0^t\dd s\,
  F_e(t-s)
  \big\langle
  j\cdot
  \mathbb E [O(\vect X^\beta_s)]
  \big\rangle
\end{align}
where $  \langle
  j\cdot
  \mathbb E [O(\vect X^\beta_s)]
  \rangle$ is the \emph{equilibrium time
  correlation} between the observation $O$ at time $s$ and the flux $j$.

@@

\subsection{The non-equilibrium response theory}

The classical response theory mainly has the following two limitations:
Firstly, one has to start from an equilibrium simulation. That means if the
external driving is no longer small ($\eps$ is relatively big), the
first order (or linear) response~\eqref{eqn:eq-lr} may not be a good approximation to
the true non-equilibrium process any more.
One possible way to improve the accuracy is to introduce the second
order response term, e.g., based on equation~\eqref{eqn:f2-0}.
Here one encounters the second limitation of the classical
response theory: it does not provide a frame work, under which
the second and even higher order responses can be easily derived.
The limitations of the classical responses theory motivate us to
develop a more general framework that allows one starting from any
non-equilibrium process (arbitrarily driven), and calculating linear responses from it.

This time, we consider the following Langevin equation:
\begin{align}\label{eqn:langevin-1b}
  \dot{\vect q} & = \nabla_{\vect p}\mh(\vect q,\vect p), \\\label{eqn:langevin-2b}
  \dot{\vect p} & =- \nabla_{\vect q}\mh(\vect q,\vect p)
  + (F_e(t)+\eps \Delta F_e(t)) \vect D(\vect q)
  - \gamma\vect p
  + \sigma\dot{\vect W},
\end{align}
such that for $\eps=0$ we get a driven system that is further
perturbed by the addition time-dependent driving force $ \Delta
F_e(t) \vect D(\vect q)$. Let $\vect X^\eps_t$ denote the associated solution process, and
$\mathcal P_\eps$ the probability measure of the trajectories
generated by driven Langevin
dynamics~\eqref{eqn:langevin-1b}--\eqref{eqn:langevin-2b}.

The time-dependent expectation value $O_0(t)$ of the path observable $O$ under the dynamics of the Langevin equation (\ref{eqn:langevin-1b})-(\ref{eqn:langevin-2b}) for $\eps=0$
can be written as
\begin{align}
  O_0( t) =
  \langle O[\vect X^0_s]\,\rangle_{0,t} = 
  \int\dd \vect u\, f(\vect u, 0) 
  \int_{\mc\{\vect u,0; t\}}
  O[\vect X_s]\, \dd\mathcal P_0[\vect X_s],
\end{align}
where 
$\langle \cdot \,\rangle_{0,t}$ is the ensemble average over all possible trajectories for $\eps=0$ in time interval $[0,t]$,
$f(\vect x, 0)$ is the initial distribution that may not be the equilibrium distribution, and
$\mc\{\vect u,0; t\}$ is the set of all admissible continuous trajectories starting from point
$\vect u$ at time $0$, and ending at some position at time $t$.

For $\eps\ge 0$ the  expectation value $O_\eps(t)$ takes the form
\begin{align}
  O_\eps(t) & =
  \int \dd\vect u f(\vect x,0)
  \int_{\mc\{\vect u,0; t\}} O[\vect X_s]\,
  \, \dd\mathcal P_\eps[\vect X_s]   \\
& = \int \dd\vect u f(\vect x,0)
  \int_{\mc\{\vect u,0; t\}} O[\vect X_s]\,
  \frac{\dd {\mathcal P}_\eps[\vect X_s]}{\dd\mathcal P_0[\vect X_s]}
  \, \dd\mathcal P_0[\vect X_s] .   
\end{align}
By using the Girsanov formula, 
\begin{align}\label{eqn:girsanov}
  \frac
  {\dd{\mathcal P_\eps}[\vect X_s]}
  {\dd\mathcal P_0[\vect X_s]}  
  =&\,
  \exp
  \big\{{\eps G[\vect X_s] - \frac{\eps^2}{2} H[\vect X_s]}
  \big\}
\end{align}
where the short notations are defined by:
\begin{align}
  G[\vect X_s]
  &= \frac{1}\sigma\int_0^t
  \Delta F_e(s)\,\vect D(\vect q_s)\cdot\dd\vect W_s \\
  H[\vect X_s]
  &=
  \frac 1{\sigma^2}\int_0^t
  \Delta F_e^2(s)\,\vect D^2(\vect q_s)\,\dd s.
\end{align}
Expansion of the right-hand-side of Eq.~\eqref{eqn:girsanov} in $\eps$, we have
\begin{align}\label{eqn:neq-response}
   O_\eps(t) -
  O_0(t)
  =
  \eps\, \big\langle O[\vect X_s]\cdot G[\vect X_s]\big\rangle_{0,t}
  +
  \frac{\eps^2}{2} \big\langle O[\vect X_s] \cdot (G^2[\vect X_s] - H[\vect X_s] )\big\rangle_{0,t}
  +
  \mo (\eps^3)
\end{align}
Some remarks concerning Eq.~\eqref{eqn:neq-response} seem in order:
\begin{itemize}
\item The first term on the right-hand-side of Eq.~\eqref{eqn:neq-response} is the linear
  response to the reference non-equilibrium process (driven by $F_e$, $\eps=0$). The next term is the second order response term. Obviously, one can easily write down higher orders of response by the Taylor expansion of the right-hand-side of Eq.~\eqref{eqn:girsanov}.
\item When the observable $O[\vect X_s]$ only depends on the end point of the trajectory, i.e.
  $O[\vect X_s] = O(\vect X_t)$, then $\big\langle O(\vect X_t)\cdot G[\vect X_s]\big\rangle_{0,t}$
  gives the well-know time correlation form of the linear response formula.
\item If one wants to calculate the same expectation value for a family of non-equilibrium perturbations $\Delta F_e(t)$ then 
one does not need to repeat the calculations of Eq.~\eqref{eqn:neq-response} for every member of the family. 
  If it is possible to
  express different $\Delta F_e$ in the same basis, then the responses must only be
  calculated for the single basis functions. Then with a linear combination of the responses on basis functions, one can derive
  the responses of the whole family. 
\item In the above derivation, we assume that the reference and the perturbed non-equilibrium processes
  have the same initial distribution (both of them may be different from the equilibrium distribution),
  but this is not a necessary. One can easily
  apply a similar reweighing approach to the initial distribution as we used it for the trajectory ensemble.
\end{itemize}

>>>>

\usepackage[fleqn]{amsmath}
\usepackage{amssymb,amsthm}
\usepackage[dvips]{graphicx}
\usepackage{color}
\usepackage{tabularx}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

\newtheorem{thm}{Theorem}

@@

\newcommand{\R}{{\mathbf R}}
\newcommand{\Q}{{\mathbf Q}}
\newcommand{\Z}{{\mathbf Z}}
\newcommand{\N}{{\mathbf N}}
\newcommand{\T}{{\mathbf T}}
\newcommand{\mh}{\mathcal H}
\newcommand{\eps}{\varepsilon}
\newcommand{\ml}{\mathcal L}
\newcommand{\mt}{\mathcal T}
\newcommand{\mo}{\mathcal O}
\newcommand{\mi}{\mathcal I}
\newcommand{\mc}{\mathcal C}
\newcommand{\proj}{\mathit\Pi}
\newcommand{\fwg}{{\mathcal A}}
\newcommand{\cA}{\mathcal A}
\newcommand{\bwg}{{\mathcal B}}
\newcommand{\bsigma}{\boldsymbol\sigma}
\newcommand{\bE}{{\mathbf E}}
\newcommand{\bP}{{\mathbf P}}
\newcommand{\one}{{\mathbf 1}}
\newcommand{\zero}{{\mathbf 0}}

\newcommand{\wrt}{with respect to }

@@

% \title{Linear response theory for core set identification}
\title{Linear response theory for a system under non-equilibrium conditions}
\author{Han Wang}
\author{Carsten Hartmann}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\affiliation{Zuse Institute Berlin (ZIB), Germany}
% \affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}

@@

\begin{abstract}
  In this paper, we firstly review standard equilibrium linear response
  theory in the context of molecular dynamics, and then generalize it to systems in nonequilibrium and under a nonequilibrium
  driving. We calculate not only the linear response term, but also
  second order terms for small perturbations to the reference dynamics. Furthermore, we outline how the novel nonequilibirum linear response formula can be used to compute optimal controls of molecular systems for cases in which one wants to steer the system to maximize a certain target expectation value.

\end{abstract}

@@

\section{Introduction}

Standard molecular dynamics simulations are dealing with systems in thermal equilibrium; in this case they are tuned to the canonial or Boltzmann distribution in the sense that either (1) if one starts from this distribution it remains invariant under the dynamics or (2) if one generates a very long trajectory it samples state space with respect this distribution, that is, every possible state of the molecular system under consideration is visited according to the probability given by it.  Obviously, this allows to compute equilibrium expectation values with respect to the  canonical distribution simply be computing long trajectories.

Often, however, one is interested in knowing about the response of the molecular system to perturbation out of equilibrium. The standard linear response formula allows to answer this question, at least partially. In the standard setting it gives us the first order of the change to an equilibrium expectation value as resulting from the non-equilibrium perturbation. Here, first order means first order in the size of the perturbation. This linear response formula has a long history of extensions and generalizations. In some sense it has become one of the cornerstones of modern statistical physics since it can be related to the fluctuation dissipation theorem (FDT) which roughly states that for appropriate systems in statistical equilibrium, the average response to small external perturbations can be calculated through the knowledge of suitable correlation functions of the unperturbed statistical system. In this sense there the standard linear response theory holds in a  very general sense \cite{Majda} as long as one assumes that the unperturbed system is in equilibrium.
%=Martin Hairer and Andrew J Majda, A simple framework to justify linear response theory, Nonlinearity 23 (2010) 909–922 doi:10.1088/0951-7715/23/4/008

There is an increasing number of articles in the literature that report on applications of molecular dynamics to nonequilibrium settings. There are many generalization to so-called nonequilibrium steady states based on the generality of the FDT, but despite its wide use, the present authors do not know of a linear response formula that applies to fully general nonequilibirium cases. We will provide such a formula for the case that the underlying dynamics can be described by Langevin dynamics. Furthermore, we will even provide a formula for the second order response of a Langevin system in non-equilibrium to a small perturbation.

Instead of applying this theory to a molecular system we will go one step further. We will outline how the novel nonequilibrium linear response formula can be used to compute the optimal control of molecular systems. In optimal control one seeks the optimal way to perturbed a molecular system such that a certain target expectation value (e.g. population of certain states) is maximized under constraints on the energy of the control. In general the control drives the molecular system under control out of equilibrium. Thus, the nonequilibrium linear response formula can be used to find the optimal correction of the present control regarding the expectation value of interest.

The outline of the article is as follows: First, we will review the derivation of the standard linear response formula for  rather general diffusion processes which include which Langevin dynamics. Next, we will show how to derive first and second order response formulas for the nonequilibrium case and how to apply these formula for the computation of optimal controls. Finally, we will validate the nonequilibrium linear response formula and its use for optimal control for simple test cases. This numerical experiments will also outline that the use of the linear response formula is imperative for numerical efficiency and allows to extend the applicability of linear response theory to stronger perturbations.

\section{Linear response}

In this section, we want to firstly give a simple and formal derivation of the linear response for equilibrium and nonequilibrium systems. To this end we consider a general It\^o stochastic differential equation of the form
\begin{equation}\label{sde}
dX_{t} = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t}\,,\quad t\ge 0\,,
\end{equation}
where $X_{t}\in\R^{d}$, $b(\cdot,\cdot)$ is a smooth time-dependent vector field, $a(\cdot)$ a smooth field of $d\times n$ matrices and $B_{t}$ is standard Brownian motion in $\R^{n}$. Here $v_{t}\in\R^{d}$ is any given driving force applied to the system, which in general may depend on $X_{t}$ and $t$; $\eps>0$ is a small parameter. We consider the equilibrium and nonequilibrium case separately.

\subsection{Small perturbations from equilibrium: Langevin dynamics}

For the equilibrium case (see, e.g., \cite{XXX,YYY,ZZZ}), we assume that the infinitesimal generator
\[
\fwg^{\eps} = \fwg_{0} + \eps\fwg_{1}\,, 
\]
with 
\[
\fwg_{0} = \frac{1}{2}aa^{T}\colon\nabla^{2} + b\cdot\nabla\,,\quad \fwg_{1}=v\cdot\nabla 
\]
of (\ref{sde}) has an isolated eigenvalue 0 corresponding to the unique invariant measure of the dynamics for $\eps=0$ and satisfies a spectral gap condition (see, e.g., \cite{stoltz2012}), so that the operator $(\fwg_{0}^{*})^{-1}\fwg_{1}^{*}$ is bounded (on an appropriate domain), where $\cA_{0}^{*}$ and $\cA_{1}^{*}$ denote the formal adjoints in $L^{2}$, e.g., $\cA_{1}^{*}\phi=-\nabla(v \phi)$.

Specifically, we are interested in the case that (\ref{sde}) has the form of a Langevin equation, in which case  
\begin{equation}\label{eqn:langevin-1}
b = \left(\nabla_{p}H, -\nabla_{q} H - \gamma\nabla_{p}H\right )^{T}\,,\quad a= (\zero,\sigma)^{T}\,,
\end{equation}
where $H\colon\R^{d}\to\R$, $d=2n$ is the Hamiltonian of the system, $\gamma=\gamma^{T}>0$ the positive semi-definite $n\times n$ friction matrix, $\sigma$ constant, satisfying $2\gamma=\beta\sigma\sigma^{T}$ for some $\beta>0$, and we have used the convention $x=(q,p)\in\R^{n}\times\R^{n}$. Under these conditions on friction and noise and some mild growth conditions on the Hamiltonian for large arguments, the unperturbed dynamics has a unique invariant measure with density 
\[
\rho_{0}(q,p) = \frac{1}{Z}e^{-\beta H(q,p)}\,,\quad  Z=\int  e^{-\beta H(q,p)}\,dqdp\,.
\] 
and the above spectral gap condition is met. Now let $f$ be any integrable phase space function and let $\rho^{\eps}=\rho^{\eps}(q,p,t)$ denote the probability density of $X_t=(Q_{t},P_{t})$, the solution to the Langevin equation (\ref{sde})--(\ref{eqn:langevin-1}), assuming that $(Q_{0},P_{0})=(q,p)$ was distributed according to the invariant density $\rho_{0}$. We define the expectation \wrt $\rho^{\eps}$ as 
\[
\bE_{\rho^{\eps}}[f] = \int f \rho^{\eps} \,dqdp\,.
\] 
A classical result (see, e.g., \cite{HairerMaijda08} and the references therein), that is usually derived using a formal expansion of the solution to the Kolmogorov forward or backward equations in powers of $\eps$, now states that 
\begin{equation}\label{lr}
\lim_{\eps\to 0}\frac{\bE_{\rho^{\eps}}[f] - \bE_{\rho^{0}}[f]}{\eps} = \bE_{\rho^{0}}[\cA_{1}\cA_{0}^{-1}(f-\bE_{\rho^{0}}[f])]\,.
\end{equation}

\subsubsection*{Green-Kubo relations} 
For the specific case of the Langevin equation (\ref{sde})--(\ref{eqn:langevin-1}), the general linear response (\ref{lr}) can be recast in form of the better known Green-Kubo formula \cite{MarconiPhysRep,Risken}. Using the formal operator identity 
\[
-\cA_{0}^{-1} g = \int_{0}^{\infty}e^{t\cA_{0}}g \,dt\,,
\]
that can shown to hold for the generator of the Langevin equation under mild conditions (formally for functions $g$ that are orthogonal to the nullspace of $\cA_{0}^{*}$, i.e. functions satisfying $\bE_{\rho_{0}}[g]=0$), we find that 
\begin{equation}\label{GreenKubo}
\lim_{\eps\to 0}\frac{\bE_{\rho^{\eps}}[f] - \bE_{\rho^{0}}[f]}{\eps} = -\beta\int_{0}^{\infty}\bE_{\rho^{0}}[J(Q_{0},P_{0}) (f(Q_{t},P_{t})-\bE_{\rho^{0}}[f])]\,,dt
\end{equation}
in terms of the dissipative flux $J=-v\cdot\nabla H$, where the expectation under the integral is taken over all realizations/path of the unperturbed Langevin equation with initial distribution $\rho^{0}(q,p,0)=\rho_{0}$. In other words: 
\[
\bE_{\rho^{\eps}}[f] \approx \bE_{\rho_{0}}[f] - \eps\beta\int_{0}^{\infty}\bE_{\rho_{0}}[J(Q_{0},P_{0}) (f(Q_{t},P_{t})-\bE_{\rho_{0}}[f])]\,dt.
\]

@@

\subsection{Nonequilibrium response theory: controlled Langevin dynamics}

The classical response theory has the limitation that the reference (equilibrium) distribution $\rho_{0}$ must be stationary under the dynamics and is unique. Moreover the validity of the linear approximation relies on a spectral gap condition that is difficult to verify in practice. In particular the perturbation argument does not provide a framework, under which the second and even higher order responses can be easily derived.

\subsubsection*{Girsanov transformation}

Here we propose an alternative (purely formal) derivation of the linear response result, that is based on a change of drift in the corresponding SDE and which allows for an easy generalization of the above linear response result to nonequilibrium systems. We will briefly review the idea of the change of drift via Girsanov transformations; for details we refer to the textbook \cite{oksendal}. Now let $x_{t},X_{t}$ be the solutions to the following stochastic differential equations:
\begin{subequations}\label{sde2}
\begin{align}
dx_{t} & = b(x_{t},t)dt + a(x_{t})dB_{t} \label{sde2-1}\\
dX_{t} & = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t} \label{sde2-2}
\end{align}
\end{subequations}
for $0\le t\le T$ and with fixed initial conditions
\[
x_{0} = X_{0} = x\,,
\]
Suppose that there exists a auxiliary stochastic process $\xi_{t}\in\R^{m}$ such that 
\begin{equation}\label{xi}
a(X_{t})\xi_{t} = v_{t}\,,
\end{equation}
The auxiliary variable $\xi$ will be called \emph{control variable}. We define  
\[
W_{t} = \eps\int_{0}^{t}\xi_{s}\,ds + B_{t}\,,\quad 0\le t\le T\,,
\]
which allows us to express (\ref{sde2-2}) by
\begin{equation}\label{sde2-3}
dX_{t} = b(X_{t},t)dt +a(X_{t})dW_{t} 
\end{equation}
It follows from the Girsanov theorem \cite[Thm.~8.6.8]{oksendal2003}, sometimes also called \emph{Cameron-Martin-Girsanov theorem} \cite{}, that $W_{t}$ is again a Brownian motion under a new probability measure that has a density \wrt the Gaussian probability measure that is generated by the Brownian motion $B_{t}$. Specifically, let $\nu$ denote the law of the Brownian motion $B_{t}$ and define a new probability measure $\mu$ on the space of continuous trajectories by 
\[
d\mu=M_{T}d\nu
\]
with 
\begin{equation}\label{likelihood}
M_{t} = \exp\left(-\eps\int_{0}^{t}\xi_{s}\cdot dB_{s} - \frac{\eps^{2}}{2}\int_{0}^{t}|\xi_{s}|^{2}ds\right)\,,\quad 0\le t\le T\,.
\end{equation}
Technical details aside, the Girsanov theorem implies that $W_{t}$  for any function $f=f(\{X_{t}\}_{0\le t\le T})$ that is integrable \wrt $\nu$, we have the identity
\[
\bE_{\nu}[f] := \int f(\{X_{t}\}_{0\le t\le T})\, d\nu = \int f(\{X_{t}\}_{0\le t\le T}) \frac{d\nu}{d\mu}\,d\mu =: \bE_{\mu}[M^{-1}_{T}f]\,.
\]
where 
\[
M_{T}^{-1} = \exp\left(\eps\int_{0}^{T}\xi_{s}\cdot dW_{s} - \frac{\eps^{2}}{2}\int_{0}^{T}|\xi_{s}|^{2}ds\right)\,.
\]
is the density of $\nu$ relative to $\mu$.

Linearization of $M_{T}^{-1}$ about $\eps=0$, assuming that the control has bounded variance, yields the alternative linear response formula 
\begin{equation}\label{lr-alt}
\lim_{\eps\to 0}\frac{\bE_{\nu}[f] + \bE_{\mu}[f]}{\eps} = \bE_{\mu}\left[f(\{X_{t}\}_{0\le t\le T})\int_{0}^{T}\xi_{s}\cdot dW_{s} \right]\,,
\end{equation}
where $X_{t}$ is the solution (\ref{sde2-3}). Note that this is the unperturbed nonequilibrium dynamics, because $W_{t}$ is a standard Brownian motion under $\mu$, and the expectation is over all realizations of (\ref{sde2-3}) starting from either any given initial condition $X_{0}=x$ or an arbitrary initial distribution. On the other hand, $X_{t}$ under $\nu$ corresponds to the perturbed dynamics, which should become clear upon comparing equations (\ref{sde2-2}) and (\ref{sde2-3}). Further note that $f$ and $W_{t}$ are not independent, hence the expectation on the right hand side of the equation is not zero in general.

\paragraph*{Remark.} A quick-and-dirty derivation of the above change-of-measure formula can be easily obtained, if the noise covariance $a(\cdot)a(\cdot)^{T}$ has full rank almost everywhere with bounded inverse. Then, using Euler's method for (\ref{sde2}), it follows that (\ref{likelihood}) is basically the likelihood ratio between the time-discrete path densities of (\ref{sde2-2}) and (\ref{sde2-1}).

\paragraph*{Remark.} By formally expanding $M^{-1}_T$ up to second order we get an analogous second order response formula: 
\begin{equation}\label{2nd}
\bE_{\nu}[f] \approx \bE_{\mu}[f] + \eps \bE_{\mu}\left[f\int_{0}^{T}\xi_{s}\cdot dW_{s} \right] +\frac{\eps^{2}}{2} \bE_{\mu}\left[f\left\{\left(\int_{0}^{T}\xi_{s}\cdot dW_{s}\right)^2 -\int_0^T |\xi_s|^2\,ds\right\} \right]
\end{equation}

\subsubsection*{Nonequilibrium Langevin equation}

We now link our previous considerations with the previous case and consider a non-equilibrium Langevin equation. Specifically, we add a non-gradient perturbation to the Langevin equation 
\begin{equation}\label{eqn:langevin-1b}
b = \left(\nabla_{p}H, -\nabla_{q} H - \gamma\nabla_{p}H  + D(q) u \right)^{T}\,,\quad a= (\zero,\sigma)^{T}\,,
\end{equation}
where $u\in\R^{n}$ is some control variable and $D(\cdot)\in\R^{n\times n}$ satisfies the Fredholm alternative $\mathrm{range}(D(\cdot))\perp \mathrm{ker}(\sigma^{T})$, where $\mathrm{ker}(\sigma^T)$ denotes the kernel of $\sigma^T$. As we are interested in small perturbation $\delta u$ in the controls $u$, it is convenient to further specify what $v$ in (\ref{sde2}) is. Specifically, we assume that $v$ is of the form 
\begin{equation}\label{eqn:langevin-1c}
v = (\zero, D(q)\,\delta u)
\end{equation}
so that equation (\ref{xi}) that determines the change of measure in terms of the auxiliary control variable $\xi$ (and thus the linear response) reads 
\[
\sigma \xi_{t} = D(Q_{t})\,\delta u_{t}\,.
\]
To be more explicit, the equation we are considering has the form
\begin{equation}\label{eqn:langevin-1d}
\begin{aligned}
dQ_t & =  \nabla_p H(Q_t,P_t)\\
dP_t & =  -\nabla_q H(Q_t,P_t) - \gamma \nabla_p H(Q_t,P_t) + D(Q_t)\,(u_t+\eps \delta u_t)+\sigma dB_t,
\end{aligned}
\end{equation}
so that $X_t$ from above now is $(Q_t,P_t)$ and $x_t=(q_t,p_t)$ from above is the solution of our last equation for $\eps=0$.
